---
title: "Prediction Assignment Writeup"
author: "Federico Roman"
date: "24 July 2015"
output: html_document
---

##Background
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset). 

##Data 

* The training data for this project are available here: 
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

* The test data are available here: 
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har. If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment. 

##Analysis
###Load packages
```{r,warning=FALSE,message=FALSE}
library(caret)
library(dplyr)
library(randomForest)
```

```{r,echo=FALSE}
setwd("~/Library/Mobile Documents/com~apple~CloudDocs/ML")
```

###Download dataset
```{r,warning=FALSE, message=FALSE}

url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
dir.create("data",showWarnings = FALSE)
path <- paste("data", "pml-training.csv", sep = "/")
  
if(!file.exists(path)) { 
  download.file(url, path, method = "curl") 
}
```

###Read data
```{r}
pmltrain <- read.csv("data/pml-training.csv")
```
###Split dataset
Split dataset into training (60%) and testing (40%)
```{r}
set.seed(1234)

# Taking 60% for the training and 40% for the test 
inTrain <- createDataPartition(y = pmltrain$classe, list = FALSE, p=0.6)
trainData <- pmltrain[inTrain,]
testData <- pmltrain[-inTrain,]
```
From now one we will use the training dataset.

###NAs value analysis
```{r}
# calculate the percentage of NAs in each column
nacol <- colSums(is.na(trainData))/nrow(trainData)
# select the columns where the percentage is greater than 80%
colsremove <- which(nacol > 0.80) 
# these are the columns that will not be considered
nacolnames <- names(nacol[nacol > 0.80]) 
nacolnames
# take a sample from the training set to workout the predictors
sampletrain <- sample_n(trainData,size=1000)
# remove columns with mostly NAs
sampletrain <- sampletrain[,-colsremove]
# remove X,username or timestamp columns  
sampletrain <- sampletrain[,-grep("X|user_name|cvtd_timestamp",names(sampletrain))]
# remove columns predictors with very small variance.
sampletrain <- sampletrain[,-nearZeroVar(sampletrain)]
```

###Random forest
List of predictors
```{r}
prednames <- names(sampletrain)
prednames2 <- prednames[-grep("classe",prednames)] 
#predictors
head(prednames2)
```
Model building
```{r}
cleanedTrainData <- trainData[,prednames]
modelFit <- randomForest(classe ~., data=cleanedTrainData, type="class")
```
Generate error estimates for "classe" variable
```{r}
classe_col <- grep("classe",names(testData))
predTest <- predict(modelFit, newdata = testData[,-classe_col], type="class")
confusionMatrix(predTest,testData$classe)
```
The model is 0.9978 accurate.

###Predict using the model
Download test data
```{r,warning=FALSE, message=FALSE}

url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
dir.create("data",showWarnings = FALSE)
path <- paste("data", "pml-testing.csv", sep = "/")
  
if(!file.exists(path)) { 
  download.file(url, path, method = "curl") 
}
```
Load the test data
```{r}
pmltest <- read.csv("data/pml-testing.csv")
```
Perform the prediction
```{r}
predplmtest <- predict(modelFit, newdata = pmltest, type="class")
print(predplmtest)
```

